Optimization

### 2020-11-24 optim/opt-talk/optimization.Rmd
First draft

## Quadratic programming

>>>>

## Quadratic programming - 1
- The problem does not get complicated if the objective function is a quadratic form.
That is, it can be written as $$Q = \frac{1}{2}x^T\Sigma x. \tag{1}$$ where $x \in \mathbb{R}^n$
and $\Sigma$ is an $n \times n$ symmetric matrix.
- Consider the problem of allocating a certain amount of money into asset classes with the objective of
minimizing the risk while still ensuring a certain return. It is equivalent to finding out what fraction 
$x_i$ of the available money should be invested in an asset 1 ≤ i ≤ N . Let the mean returns of the $N$ assets be
$r_1 , . . . , r_N$ and let $\Sigma$ be the covariance matrix of the returns.
- The problem then is to minimize the quantity $Q$ in equation (1) subject to
  - $x_1r_1 + \cdots + x_nr_n \ge r$, where $r$ is that minimum return desired.
  - $x_1 + \cdots + x_n = 1$, that is, the entire money is invested and
  - $x_1, \ldots, x_n \ge 0$, that is, there is not shorting.

## Quadratic programming - 2
- We simulate the returns and the variance-covariance matrix of the returns of
a portfolio of $N = 5$ assets.
- Generating a random covariance matrix is a little involved process for it has to be positive semidefinite.
We use the fact that if Q is an orthogonal matrix and D a diagonal matrix with all non-negative entries
then QT DQ is positive semi-definite. We can extract an orthogonal matrix from any matrix by its QR
decomposition.

```{r echo=TRUE}
set.seed(19471508)
N <- 5 # number of assets
mu <- rnorm(n = N, mean = 6, sd = 2) # mean returns

# Randomly generated covariance matrix.
M <- matrix(data = rnorm(N*N), nrow = N, ncol = N)
Q <- qr.Q(qr(M))
C <- Q %*% diag(seq(from = 1, to = 5)) %*% t(Q) # The variance-covariance matrix.
```

## Quadratic programming - 3
- We solve the quadratic program.

```{r echo=TRUE}
library(quadprog)
k <- rep(1, N)
r <- 6.5 # Minimum return
b <- c(r, 1, -1, rep(0, 5))
A <- rbind(mu, k, -k, diag(rep(1, N)))
qp.results <- solve.QP(Dmat = C, dvec = rep(0, N), Amat = t(A), bvec = b)
s <- round(qp.results$solution, 4)
names(s) <- paste(sep = "", "asset", seq(from = 1, to = N))
cat("The optimal asset allocation is:\n")
print(s)
cat(paste(sep = "", "The risk of the portfolio is ", round(qp.results$value, 4), ".\n"))
```

## Closing remarks
- You might feel that optimization is too technical and dry. Yet,
  - Several problems in physics and engineering are cast as (nonlinear) optimization problems.
  - Optimization is at the heart of many AI/ML problems. You may have heard of 
  terms like gradient descent,conjugate gradient descent, batch gradient descent, 
  XGBOOST, ADABOOST etc. Every one of them is an optimization algorithm.
  - Even traditional statistical problems like ordinary least square regression
  and logistic regression are solved with an optimizer. You **minimize** an
  error function in each of them.
- Linear optimization might not be glamorous but the simplex method is the 
commercially most successful algorithm. Some examples:
  - How to transport $m$ items from $n$ factories to $l$ warehouses in the 
  *cheapest* way?
  - How to match $m$ people to $n$ jobs to get the *maximal* matching?
  - How to build a schedule of buses so that they are idle for the *least* time?
  - How to design a tour of $n$ cities that is the *cheapest*? (This problem
  is computationally very expensive.)
